---
title: "DE analysis with edgeR and contrast"
author: "Leticia"
date: "02/08/2021"
output: html_document
---

#######################################################################################################################################################################################
#Load libraries
```{r setup, include=FALSE}
library(tidyverse)
library(edgeR)
library(limma)
library(tidyr)
library(dplyr)
library(FactoMineR)
library(ggplot2)
library(gplots)
library(marray)
library(affy)
library(CLL)
library(Biobase)
library(dendroextras)
library(RColorBrewer)
library(biomaRt)
source("/path/diffexpr4.R")
```

# Set the workinf directory
```{r  workingdirectory, warning=FALSE, message=FALSE}
setwd("/path/DE_Analysis/")
```

#####################################################################################################################################################################
We can use this script for DE analysis just right after the one for filter and normalization
#For DE analysis we need to use Raw data
#####################################################################################################################################################################
#### Load data, counts and metadata for the cmb from cmbBase
################################################################################
```{r  loadData, warning=FALSE, message=FALSE}
f.counts <- as.matrix(read.table( "/path/filter_raw.counts", sep = "\t", header = TRUE, row.names = 1, as.is=T))
class(f.counts)
dim(f.counts)
Metadata
ss <-read.csv("/path/metadata.csv", sep = ",")
dim(ss)
#sanitary check to be sure that columns in counts as same as row in ss
identical(as.character(ss$sample), colnames(f.counts))
```

#############################################
############################################# 
#############################################
# GLM and CONTRAST
### We are using only one design, GLM with contrats
##### contrast1 for the the comparisons which are the same but the other way around
```{r model&contrast, message=FALSE, warning=FALSE}
#rs <- paste(sstumor$Type, sstumor$N_Reads, sep = ".")
#nreads <- factor(sstumor$N_Reads)

#take and create teh models using the metadata column type
rs <- factor(ss$Type )# levels = unique(rs))
design <- model.matrix(~ 0 + rs ) #fitting an interaction model
colnames(design) <- levels(rs)
head(design)
#second design in case we want to make other comparisons
design2 <- model.matrix(~ 0 + ss$group ) #fitting an interaction model 
colnames(design2) <- levels(ss$group)
head(design2)

#prepare the contrast== comparisons to make, usign the names in column type, this is an example
contrast <- makeContrasts(Tm_Ctrl=Tumor-Nv, Crtl_Tm=Nv-Tumor, levels = design)

#we can have more than one type of comparisons, we need also a 2nd design for that
contrast2 <- makeContrasts(TPLL_Memory=TPLL-Memory, TPLL_EF=TPLL-EF, TPLL_Nv=TPLL-Naive, levels = design2)

```
## Creating DGEList elements, Normalization and Dispersion
#Para hacer el exact text tengo que ponerle los grupos en incluirlo al crear la DEGList
## Cuidado con el calculo del adispersion, varian los resultados, sobre todo pVal
```{r DGElist, message=FALSE, warning=FALSE}
#group <- sstumor$Type
D <- DGEList(counts = as.matrix(cmb))
Dnorm <- calcNormFactors( D )
Dnorm$samples
plotMDS(D)
lin.norm.factors <- (sum(D$samples$lib.size)/nrow(D$samples))/ D$samples$lib.size
lin.norm.factors
##Calculating common and tagwise dispersion in one go, using the estimateDisp function that exploits the Cox-Reid profile-adjusted likelihood (CR) method in estimating dispersions with a specified design matrix
Dt <- estimateDisp(Dnorm, design)
#Dc <- estimateGLMCommonDisp( Dnorm, design) #Calculating common dispersion
#Dt <- estimateGLMTagwiseDisp( Dc, design, prior.df=2 ) #Calculation Tagwise dispersion
Dt.fit <- glmFit(Dt, design)
plotBCV(Dt)
Dt

## This step is below for the contrast
#Dt.lrt <- glmLRT(Dt.fit)
```


Now Using CONTRAST and GLM_LRT to create the table and save the data
###########################################################################3
Here we start to use the contrast/ comparisons that we prepare before with the model design, Type
###### Contrats 1
```{r contrast_tablecreate, message=FALSE, warning=FALSE}
#table create using the name of the contrast
Dt.lrt1 <- glmLRT(Dt.fit, contrast = contrast2[,"TPLL_EF"])
topTags(Dt.lrt1)

Dt.lrt.table1 <- topTags( Dt.lrt1, n=nrow(Dt$counts) )$table
write.table(Dt.lrt.table1, file="name.txt", sep="\t")
write.csv(Dt.lrt.table1, file="name.csv")
```

####### Data formatting
```{r dataformatting, message=FALSE, warning=FALSE}
#sorting by FDR
sort_fdr <- sort(Dt.lrt.table1$FDR)<0.05
pval <- topTags(Dt.lrt1, n=nrow(sort_fdr), adjust.method = "BH", sort.by = "PValue")$table
dim(pval)
#take only the significant (FDR)
fdr_pass <- pval[(pval$FDR)<0.05,] # first cutt-off, FDR<0.05
dim(fdr_pass)
#from the significant take only the ones with logFC >1 (FC>2)
lgfc_pass <- fdr_pass[which(abs(fdr_pass$logFC)>1),] # Second cutt-off, log2FC(abs=1) or FC(abs=2)
lgfc_pass$FC <- 2^lgfc_pass$logFC
dim(lgfc_pass)
#save the data
write.table(lgfc_pass, file="Significant_listgenes.txt", row.names = TRUE, col.names = NA, quote=FALSE, sep="\t")

#Normalized counts
tmm <- cpm(Dnorm, normalized.lib.sizes = TRUE)
#tmm_gp <- cpmByGroup(y, group = NULL, dispersion = NULL, ...)
#Nrpkm <- rpkm(Dnorm, gene.length = NULL, normalized.lib.sizes = NULL, log = FALSE )
#Nrpkm_gp <-rpkmByGroup(y, group = NULL, gene.length = NULL, dispersion = NULL, ...)

#we take the rows (significants genes) go to the Norm matrix and take the counts of the significant genes
sign_tmm <- tmm[c(rownames(lgfc_pass)),]
head(sign_tmm)  #####Careful is a Matrix
dim(sign_tmm)
sign_tmm <- as.data.frame(sign_tmm)
#sign_tmm$mean_CD <- rowMeans(sign_tmm[,5:9])
#sign_tmm$mean_TPLL <- rowMeans(sign_tmm[,35:55])
head(sign_tmm)
dim(sign_tmm)

#save the counts of the significant genes 
write.table(sign_tmm, file="Sign_cpm_.txt", row.names = TRUE, col.names = NA, quote=FALSE, sep="\t")
sign_DE_tmm <- cbind(lgfc_pass, sign_tmm)
dim(sign_DE_tmm)
head(sign_DE_tmm)
write.table(sign_DE_tmm, file="Significant_counts.txt", row.names = TRUE, col.names = NA, quote=FALSE, sep="\t")
```


################Curve for the Pvalues 
```{r pvalCurve, message=FALSE, warning=FALSE}
pdf("FDR_.pdf")
plot(sort(Dt.lrt.table1$FDR))
abline(c(0,1/nrow(Dt.lrt.table1)), col="blue")
dev.off()
```
#####To make the Boxplots
```{r boxplots, message=FALSE, warning=FALSE}
number_of_plots <- 100
top_genes <- row.names(Dt.lrt.table1)[1:number_of_plots]
pdf("boxplots_.pdf")
for (i in 1:length(top_genes))
{
 boxplot(as.numeric(log(cpm(Dnorm[which(rownames(Dnorm)==top_genes[i]), ])))~as.factor(ss_miR$Type), main=paste(top_genes[i],"FDR", Dt.lrt.table1$FDR[which(rownames(Dt.lrt.table1)==top_genes[i])], sep= " "))
}
dev.off()
```

#############################################################3 
Next contrast/comparison that we prepare before with the model design, Type
###### Contrats2  
```{r contrast_tablecreate, message=FALSE, warning=FALSE}
Dt.lrt2 <- glmLRT(Dt.fit, contrast = contrast2[,"TPLL_Nv"])
topTags(Dt.lrt2)
Dt.lrt.table2 <- topTags( Dt.lrt2, n=nrow(Dt$counts) )$table
write.table(Dt.lrt.table2, file="name.txt", sep="\t")
write.csv(Dt.lrt.table2, file="name.csv")
```

####### Data formatting
```{r dataformatting, message=FALSE, warning=FALSE}
sort_fdr <- sort(Dt.lrt.table2$FDR)<0.05
pval <- topTags(Dt.lrt2, n=nrow(sort_fdr), adjust.method = "BH", sort.by = "PValue")$table
dim(pval)
fdr_pass <- pval[(pval$FDR)<0.05,] # first cutt-off, FDR<0.05
dim(fdr_pass)
lgfc_pass <- fdr_pass[which(abs(fdr_pass$logFC)>1),] # Second cutt-off, log2FC(abs=1) or FC(abs=2)
lgfc_pass$FC <- 2^lgfc_pass$logFC
dim(lgfc_pass)
write.table(lgfc_pass, file="Significant_geneslist.txt", row.names = TRUE, col.names = NA, quote=FALSE, sep="\t")
tmm <- cpm(Dnorm, normalized.lib.sizes = TRUE)
#tmm_gp <- cpmByGroup(y, group = NULL, dispersion = NULL, ...)
#Nrpkm <- rpkm(Dnorm, gene.length = NULL, normalized.lib.sizes = NULL, log = FALSE )
#Nrpkm_gp <-rpkmByGroup(y, group = NULL, gene.length = NULL, dispersion = NULL, ...)
sign_tmm <- tmm[c(rownames(lgfc_pass)),]
head(sign_tmm)  #####Careful is a Matrix
dim(sign_tmm)
sign_tmm <- as.data.frame(sign_tmm)
#sign_tmm$mean_CD <- rowMeans(sign_tmm[,5:9])
#sign_tmm$mean_TPLL <- rowMeans(sign_tmm[,35:55])
head(sign_tmm)
dim(sign_tmm)
write.table(sign_tmm, file="Sign_cpm_genescounts.txt", row.names = TRUE, col.names = NA, quote=FALSE, sep="\t")
sign_DE_tmm <- cbind(lgfc_pass, sign_tmm)
dim(sign_DE_tmm)
head(sign_DE_tmm)
write.table(sign_DE_tmm, file="Significant_genescounts.txt", row.names = TRUE, col.names = NA, quote=FALSE, sep="\t")
```


################Curve for the Pvalues 
```{r pvalCurve, message=FALSE, warning=FALSE}
pdf("FDR_.pdf")
plot(sort(Dt.lrt.table2$FDR))
abline(c(0,1/nrow(Dt.lrt.table2)), col="blue")
dev.off()
```
#####To make the Boxplots, this part is t
```{r boxplots, message=FALSE, warning=FALSE}
number_of_plots <- 100
top_genes <- row.names(Dt.lrt.table2)[1:number_of_plots]
pdf("boxplots.pdf")
for (i in 1:length(top_genes))
{
 boxplot(as.numeric(log(cpm(Dnorm[which(rownames(Dnorm)==top_genes[i]), ])))~as.factor(ss_miR$Type), main=paste(top_genes[i],"FDR", Dt.lrt.table2$FDR[which(rownames(Dt.lrt.table2)==top_genes[i])], sep= " "))
}
dev.off()
```


####
And so on and so on, 
We can add and create as much contrast as we want 

















# System info
```{r}
sessionInfo()
Sys.info()
```