---
title: "Script for filter, claculate Norm and plot heatmaps"
author: "Leticia G Leon"
date: "05/08/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(gplots)
library(edgeR)
library(DESeq2)
library(FactoMineR)
library(biomaRt)
library(Rsubread)
library(Biobase)
library(getopt)
library(vsn) 
library(BiocParallel)
library(GenomicAlignments)
library(GenomicFeatures)
library(Rsamtools)
library(pheatmap)
library(RColorBrewer)
library(PoiClaClu)
library(genefilter)
library(ggbeeswarm)
library(apeglm)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(ReportingTools)
library(Gviz)
library(sva)
library(RUVSeq)
library(hexbin)
source("/path/diffexpr4.R")
```

### Setting WD
```{r setwd, warning=FALSE, message=FALSE}
setwd("/path/working_dir/")
```

Load data
#we load the filter counts (without low counts data)
```{r loadData, warning=FALSE, message=FALSE}
#load the filter data
f.counts <- read.delim("/path to filter data.txt", header = T, sep = "\t", as.is = T, row.names = 1)
#this is in case we did not remove the columns at the beginning
f.counts <- f.counts[,-c(1:4)] 

head(f.counts)
#laod metadata
ss <- read.csv("/path/metadata.csv",   sep = ";")
head(ss)


```

Creating a DESeq2 object, linking with counts and metadata
```{r DESeq2object, warning=FALSE, message=FALSE}
ddsMatN <- DESeqDataSetFromMatrix(countData = f.counts[,-1], 
                                 colData = ss,
                                 design = ~ TypeLession ) #here use and var column in the ss that you will use as design model
```

##Calculate rlog (this is a type of Normalization, very good for visualization and plotting)
```{r rlog, warning=FALSE, message=FALSE}
ddsN_rlog <- rlog(ddsMatN, blind = T,   fitType = "mean") #blind is to make it completely unsupervised, no looking at the conditions or not 
head(assay(ddsN_rlog)) #this is because the data is store at the assay slot (R)
```

Represent rlog versus a typical log2 transformation of the data
sanitary check
```{r plot_rlog,  warning=FALSE, message=FALSE}
ddsMatN <- estimateSizeFactors(ddsMatN)

df <- bind_rows(
  as_data_frame(log2(counts(ddsMatN, normalized=T)[,1:2]+1)) %>%
    mutate(transformation="log2(x+1)"),
  #as_data_frame(assay(dds_vst)[, 1:2]) %>% dplyr::mutate(transformation = "vst"),
  as_data_frame(assay(ddsN_rlog)[,1:2])%>% mutate(transformation="rlog"))

colnames(df)[1:2] <- c("x", "y")

ggplot(df, aes(x=x, y=y)) + 
  geom_hex(bins = 80) +
  coord_fixed() +
  facet_grid(. ~ transformation)
  
```
########################################################################################################################################################################################################################################################################################################################
# Data  checking after rlog Normalization, using rlog(DESeq2) tools
#### Now we check the distances using the rlog matrix
```{r distance, warning=FALSE, message=FALSE}
#withNormals
sampleDists <- dist(t(assay(ddsN_rlog)))

#sampleDist
sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste(ddsN_rlog$Patient.ID, ddsN_rlog$TypeLession, sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```

#### Calculate and see rlog data as poisson distribution
```{r, warning=FALSE, message=FALSE}
#withNormals
poisd <- PoissonDistance(t(counts(ddsMatN)))

samplePoisDistMatrix <- as.matrix( poisd$dd )
rownames(samplePoisDistMatrix) <- paste( ddsMatN$Patient.ID, ddsMatN$TypeLession, sep=" - " )
colnames(samplePoisDistMatrix) <- NULL
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         col = colors)
```

#### PCA
```{r, warning=FALSE, message=FALSE}
#withNormals
plotPCA(ddsN_rlog, intgroup = c("TypeLession", "Grade_dysplasia"))


# Using day and type column information for the PCAs
pcaData <- plotPCA(ddsN_rlog, intgroup = c( "TypeLession", "Grade_dysplasia"), returnData = TRUE)  
pcaData

percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(x = PC1, y = PC2, color = TypeLession, shape = Grade_dysplasia)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed()

## Here with day and group the PCAs
pcaDatag <- plotPCA(ddsN_rlog, intgroup = c( "Grade_dysplasia", "TypeLession"), returnData = TRUE)
pcaDatag

percentVar <- round(100 * attr(pcaData, "percentVar")) # Is giving how much this Variab contribute in the PCA in 100% 

ggplot(pcaDatag, aes(x = PC1, y = PC2, color = Grade_dysplasia, shape = TypeLession)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +  # How is going to appear in the plot
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed()

```

##################################################################################################################################

Extra-visualization, using zscores and other methods

##################################################################################################################################

####Load Norm data
For rlog Norm data, don't need to load anything, just use your recently rlog transformed data 
But we can load other Norm data (rpkm, tpm or fpkm)
```{r loadNormdata, message=FALSE, warning=FALSE}
org.rlog <- read.delim("/path/data_rlogNorm.txt", header = T, sep = "\t", as.is = T, row.names = 1)
head(org.rlog)
org.rpkm <- read.delim("//rpkm_data.txt", header = T, sep = "\t", as.is = T, row.names = 1)
head(org.rpkm)
#metadata again, could be hte same one as before
ss_ <- read.csv("/path/ss_norm_.csv",   sep = ";")
head(ss_)

```

We can even use a subset of data, selecting specific genes
Selection of the genes MACROD2, WWOX, FHIT, PRKN, NOTCH2, etc etc
```{r geneselct, warning=FALSE, message=FALSE}
 five.org.rlog<- org.rlog[c("MACROD2", "WWOX", "FHIT", "PRKN", "NOTCH2"),]
head(five.org.rlog)
five.org.rpkm <- org.rpkm[c("MACROD2", "WWOX", "FHIT", "PRKN", "NOTCH2"),]
head(five.org.rpkm)

```

Make a unsupervised heatmap, with the entire matrix and the selected genes 
```{r heatmpa1, message=FALSE, warning=FALSE}
heatmap(as.matrix(org.rlog))
heatmap(as.matrix(org.rpkm))
heatmap(as.matrix(five.org.rlog))
heatmap(as.matrix(five.org.rpkm))
```
same as before but using another function, heatmaps are slightly different
```{rheatmap2,  message=FALSE, warning=FALSE}
heatmap.2(as.matrix(org.rlog), trace = "none")
heatmap.2(as.matrix(org.rpkm), trace = "none")
heatmap.2(as.matrix(five.org.rlog), trace = "none")
heatmap.2(as.matrix(five.org.rpkm), trace = "none")
```

Calculo de Zscore 
```{r Zscore, message=FALSE, warning=FALSE}
m_all <- rowMeans(org.rlog)
sd_all <- rowSds(org.rlog)
for(i in 1:nrow(org.rlog)) org.rlog[i,] <- (org.rlog[i,] -m_all[i])/sd_all[i]
```
Now we make the same heatmap usign the Zscores values
```{r heatmZscore, message=FALSE, warning=FALSE}
heatmap.2(as.matrix(org.rlog), trace = "none")
```

More elaborate heatmaps and save as pdf
```{r niceheatmap1, message=FALSE, warning=FALSE}
#splmat <- data.matrix(countsup)
#splnames <- rownames(splmat)
pdf("Clustering data.pdf")
par(mfrow = c(1,1))
my_palette <- colorRampPalette(c("darkblue", "grey","red"))(n = 299)
#col_breaks = c(seq(-6.0,-0.2,length=100),seq(-0.1,0.1,length=100), seq(0.2,6.0,length=100))
#png("tpll_zscore.png",width = 5*900, height = 5*900, res = 600, pointsize = 9) 
heatmap.2(as.matrix(org.rlog),
  #cellnote = splmat,  # same data set for cell labels
  Colv = T,
  Rowv = T,
  scale = "row",
  na.rm = TRUE,
  dendrogram = "row",
  main = "Clustering Data", # heat map title
  #notecol="black",      # change font color of cell labels to black
  density.info="none",  # turns off density plot inside color legend
  trace="none",         # turns off trace lines inside the heat map
  margins =c(10, 10),     # widens margins around plot
  col=my_palette,       # use on color palette defined earlier
  key = TRUE)
  #breaks=col_breaks)    # enable color transition at specified limits 
 
dev.off()
my_group = as.numeric(as.factor(substr(colnames(org.rlog), 1, 1)))
my_col=brewer.pal(5,"Blues")[my_group]
heatmap(as.matrix(org.rlog), scale = "row", col=heat.colors(256), main="Clustering data", ColSideColors = my_col, margins = c(10,10))
```

Clustering using different distance and hclust for columns and rows
```{r cluster1, message=FALSE, warning=FALSE}
pdf("Spearman-row-Pearson-col_c-Ward_.pdf")
# with method = ward.D2
#my_palette <- colorRampPalette(c("blue","grey", "red"))(n = 299)
hr2 <- hclust(as.dist(1-cor(t(as.matrix(org.rlog)), method="spearman")), method="ward.D2") # Clusters rows by Spearman correlation as distance method, with ward.D2
hc2 <- hclust(as.dist(1-cor(as.matrix(org.rlog), method="pearson")), method="ward.D2") # Clusters columns by Spearman correlation as distance method, with ward.D2
heatmap.2(as.matrix(org.rlog), Rowv=as.dendrogram(hr2), Colv=as.dendrogram(hc2), scale = "row", trace = "none", margins =c(10, 10), col = my_palette)
dev.off()
```

```{r cluster2, message=FALSE, warning=FALSE}
pdf("Spearman-col-Pearson-row_c-Ward_.pdf")
# with method = ward.D2
#my_palette <- colorRampPalette(c("blue","grey", "red"))(n = 299)
hr2 <- hclust(as.dist(1-cor(t(as.matrix(org.rlog)), method="pearson")), method="ward.D2") # Clusters rows by Spearman correlation as distance method, with ward.D2
hc2 <- hclust(as.dist(1-cor(as.matrix(org.rlog), method="spearman")), method="ward.D2") # Clusters columns by Spearman correlation as distance method, with ward.D2
heatmap.2(as.matrix(org.rlog), Rowv=as.dendrogram(hr2), Colv=as.dendrogram(hc2), scale = "row", trace = "none", margins =c(10, 10), col = my_palette)
dev.off()
```


##################################################################################################################################
# This part is a repetition adapted for the small matrix of the genes previously selected
Calculate zscore
```{r message=FALSE, warning=FALSE}
m_all <- rowMeans(five.org.rlog)
sd_all <- rowSds(five.org.rlog)
for(i in 1:nrow(five.org.rlog)) five.org.rlog[i,] <- (five.org.rlog[i,] -m_all[i])/sd_all[i]
```
```{r message=FALSE, warning=FALSE}
heatmap.2(as.matrix(five.org.rlog), trace = "none")
```
More elaborate heatmaps and save as pdf
```{r message=FALSE, warning=FALSE}
#splmat <- data.matrix(countsup)
#splnames <- rownames(splmat)
pdf("Fivegenesselected_.pdf")
#par(mfrow = c(1,1))
my_palette <- colorRampPalette(c("darkblue", "grey","red"))(n = 299)
#col_breaks = c(seq(-6.0,-0.2,length=100),seq(-0.1,0.1,length=100), seq(0.2,6.0,length=100))
#png("tpll_zscore.png",width = 5*900, height = 5*900, res = 600, pointsize = 9) 
heatmap.2(as.matrix(five.org.rlog),
  #cellnote = splmat,  # same data set for cell labels
  Colv = TRUE,
  Rowv = TRUE,
  scale = "row",
  na.rm = TRUE,
  dendrogram = "both",
  main = "zscore_5genes", # heat map title
  #notecol="black",      # change font color of cell labels to black
  density.info="none",  # turns off density plot inside color legend
  trace="none",         # turns off trace lines inside the heat map
  margins =c(10, 10),     # widens margins around plot
  col=my_palette,       # use on color palette defined earlier
  key = TRUE)
  #breaks=col_breaks)    # enable color transition at specified limits 
 
dev.off()
my_group = as.numeric(as.factor(substr(colnames(five.org.rlog), 1, 1)))
my_col=brewer.pal(5,"Blues")[my_group]
heatmap(as.matrix(five.org.rlog), scale = "row", col=heat.colors(256), main="5genes", ColSideColors = my_col, margins = c(10,10))
```

Clustering using different distance and hclust for columns and rows
```{r message=FALSE, warning=FALSE}
pdf("Spearman-row-Pearson-col_c-Ward_5genes_.pdf")
# with method = ward.D2
#my_palette <- colorRampPalette(c("blue","grey", "red"))(n = 299)
hr2 <- hclust(as.dist(1-cor(t(as.matrix(five.org.rlog)), method="spearman")), method="ward.D2") # Clusters rows by Spearman correlation as distance method, with ward.D2
hc2 <- hclust(as.dist(1-cor(as.matrix(five.org.rlog), method="pearson")), method="ward.D2") # Clusters columns by Spearman correlation as distance method, with ward.D2
heatmap.2(as.matrix(five.org.rlog), Rowv=as.dendrogram(hr2), Colv=as.dendrogram(hc2), scale = "row", trace = "none", margins =c(10, 10), col = my_palette)
dev.off()
```

```{r message=FALSE, warning=FALSE}
pdf("Spearman-col-Pearson-row_c-Ward_5genes_.pdf")
# with method = ward.D2
#my_palette <- colorRampPalette(c("blue","grey", "red"))(n = 299)
hr2 <- hclust(as.dist(1-cor(t(as.matrix(five.org.rlog)), method="pearson")), method="ward.D2") # Clusters rows by Spearman correlation as distance method, with ward.D2
hc2 <- hclust(as.dist(1-cor(as.matrix(five.org.rlog), method="spearman")), method="ward.D2") # Clusters columns by Spearman correlation as distance method, with ward.D2
heatmap.2(as.matrix(five.org.rlog), Rowv=as.dendrogram(hr2), Colv=as.dendrogram(hc2), scale = "row", trace = "none", margins =c(10, 10), col = my_palette)
dev.off()
```

plot distribution
```{r message=FALSE, warning=FALSE}
hist(org.rlog)
```

#those are Norm data, no sirve apra esto
```{r message=FALSE, warning=FALSE}
stable.org.rlog<- org.rlog[c("ACTB", "GAPDH"),]
head(stable.org.rlog)
```




########### The End
```{r}
sessionInfo()
Sys.info()
```