---
title: "MCP_counter and EPIC using Immunodeconv "
author: "Leticia G Leon"
date: "1/10/2021"
output: html_document
---

This script is meant to run MCP_counter and EPIC using immunedeconv (an R package for unified access to computational methods for estimating immune cell fractions from bulk RNA sequencing data).
Please follow the instructions in https://icbi-lab.github.io/immunedeconv/
To install it, before you run this code.

#```{r setup, include=FALSE}
#done, installed
library(devtools)
install_github("ebecht/MCPcounter",ref="master", subdir="Source")
#```

#```{r}
#Done, installed 
install.packages("remotes")
#library(remotes)
remotes::install_github("icbi-lab/immunedeconv")
#```

Example of using Immunedeconv, a library that is using MCPcount as one of the methods to calculate the stromal infiltration
Libraries
```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(immunedeconv)
library(tibble)
library(MCPcounter)
```

```{r load, message=FALSE, warning=FALSE}
immunedeconv::dataset_racle
knitr::kable(dataset_racle$expr_mat[1:5, ])
```

For immmunedeconv the input must be tpm-Nomr or Not log-transformed
```{r data, message=FALSE, warning=FALSE}
#for MCP_counter there is not need of Norm, but we can use TPM/RPKM is we want

C5_mcp<- as.matrix(read.delim("/Users/path/to/countfile/name.counts", row.names = 1))
class(C5_mcp)
#head(C5_mcp)
dim(C5_mcp)
#head(knitr::kable(C5_mcp, digits = 2))

```

### New part to remover the outliers and make the tmp Norm and see how it looks

Removing outliers before we Norm
```{r clean, message=FALSE, warning=FALSE}
#We need to remove columns (33, 54, 68)
C5_mcp <- C5_mcp[,-c(33, 54, 68)]
```
```{r df, message=FALSE, warning=FALSE}
C5_mcp <- as.data.frame(C5_mcp)
```

Epic Needs the data Normalized RPKM or TPM
Here we calculate tpm to use in both to test the results in both methods
```{r tpmGoodfunc, warning=FALSE, message=FALSE}
countToTpm <- function(cuentas, effLen)
{
    rate <- log(cuentas) - log(effLen)
    denom <- log(sum(exp(rate)))
    exp(rate - denom + log(1e6))
}

countToFpkm <- function(counts, effLen)
{
    N <- sum(counts)
    exp( log(counts) + log(1e9) - log(effLen) - log(N) )
}

fpkmToTpm <- function(fpkm)
{
    exp(log(fpkm) - log(sum(fpkm)) + log(1e6))
}

countToEffCounts <- function(counts, len, effLen)
{
    counts * (len / effLen)
}
```
```{r tpmGood,warning=FALSE, message=FALSE}
#countDf$effLength <- countDf$length - 203.7 + 1
tpm_c5 <- with(C5_mcp, countToTpm(C5_mcp[,-1], C5_mcp$Length))
```
####Save data tpm
```{r savetpm, warning=FALSE, message=FALSE}
write.table(tpm_c5, "/Users/path/ to / tmpNorm/rnaNOrm_tpm_lenght.txt", sep = "\t" , row.names=T, col.names=T, quote=F)
```

```{r methods, message=FALSE, warning=FALSE}
deconvolution_methods
```

how to use it
```{r howto, warning=FALSE, message=FALSE}
#resC5_epic = deconvolute(C5_mcp[,-1], "epic" , tumor = TRUE)
#resC5_mcp = deconvolute(C5_mcp[,-1], "mcp_counter")# , tumor = TRUE)
```
With Normalized data 
```{r runourdata, message=FALSE, warning=FALSE}
resC5_epic = deconvolute(tpm_c5[,-1], "epic" , tumor = TRUE)
resC5_mcp = deconvolute(tpm_c5[,-1], "mcp_counter")# , tumor = TRUE)
```

Plotting
```{r plotMCP, message=FALSE, warning=FALSE}
resC5_mcp %>%
  gather(sample, fraction, -cell_type) %>%
  # plot as stacked bar chart
  ggplot(aes(x=sample, y=fraction, fill=cell_type)) +
    geom_bar(stat='identity') +
    coord_flip() +
    scale_fill_brewer(palette="Paired") +
    scale_x_discrete(limits = rev(levels(resC5_mcp)))
```
for EPIC
```{r plotEpic, message=FALSE, warning=FALSE}
resC5_epic %>%
  gather(sample, fraction, -cell_type) %>%
  # plot as stacked bar chart
  ggplot(aes(x=sample, y=fraction, fill=cell_type)) +
    geom_bar(stat='identity') +
    coord_flip() +
    scale_fill_brewer(palette="Paired") +
    scale_x_discrete(limits = rev(levels(resC5_epic)))
```





Scatter plots for mcp_counter
```{r scaterMCP, message=FALSE, warning=FALSE}
resC5_mcp %>%
  gather(sample, score, -cell_type) %>%
  ggplot(aes(x=sample, y=score, color=cell_type)) +
    geom_point(size=4) +
    facet_wrap(~cell_type, scales="free_x", ncol=3) +
    scale_color_brewer(palette="Paired", guide="none") +
    coord_flip() +  # this is to create teh plot in horizontal, we have too many samples for that
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Scatter plots for epic
```{r scaterEpic ,message=FALSE, warning=FALSE}
resC5_epic %>%
  gather(sample, score, -cell_type) %>%
  ggplot(aes(x=sample, y=score, color=cell_type)) +
    geom_point(size=4) +
    facet_wrap(~cell_type, scales="free_x", ncol=3) +
    scale_color_brewer(palette="Paired", guide="none") +
    coord_flip() +  # this is to create teh plot in horizontal, we have too many samples for that
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```








########### The End
```{r}
sessionInfo()
Sys.info()
```